Toward theories of deliberated preferences

### 2020-08-19 simple.tex
Various suppl discussions; a sketch of a conclusion. Hidden appendices.

\section{Motivation} 
\TODO{“Before asking whether agents want to revise their choices or not, both rival theories should be proposed.” Guala 2000. My approach can be qualified as a debiasing technique (Mitchell, Libertarian paternalism is an oxymoron. Nw UL Rev 99:1245–1817, 2005). Mitchell observes that “in many of those surveys that governments use in order to inform social policy there are not enough resources (time and money) in order to reach the stage where preferences are not labile” (exact quote?).}

It has escaped nobody’s attention that 
studying reflective preferences, in supplement to shallow preferences, is an important endeavor for economics: we want to recommend to individual decision makers; we want to adopt optimization criteria for society, and this ought not to merely faithfully reproduce unreflectful behavior. This is one important difference between the normative and the (inappropriately called, as will be seen) “descriptive” approach in economics. As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow preferences are not sufficient to inform normative economics.
Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures.

@@

\subsection{Existing approaches for capturing reflective preferences}
\subsubsection{Laundering preferences}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered relevant.

This creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known. As a result, a theory postulating constraints on individual’s behavior on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. 
That this is an important methodological problem is sometimes recognized by authors adopting this approach. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making. \TODO{Comment about akrasia: we may want to correct for features that the individual herself can’t correct.}

@@

Different analysts, using different methods and different axioms as their favorite conception of rational decision-making, will obtain possibly contradictory conclusions, “elicited” from the decision-makers who hire them. Some analysts admit this possibility and do not consider it to be a problem.
For example, \citet[p.\ 214]{roy_comparison_1995} write: “In our view, these inevitable disagreements do not imply that decision-aid is useless but simply that a single problem may have several valid responses. Given that two different decision-aid models cannot be implemented in the same decision process, the decision-maker must be conscious of the qualitative choices implied by the different models – often conveying the analysts’ own ethical choices – before coming to personal conclusions on the choice to be made. In this domain, the many different approaches reflect in our view the complexity of the researcher's task much more than a scientific weakness.”
%Some analysts claim that the value of such interactions between decision-makers and analysts is the fact that it made decision-makers think hard about the problem, and not the resulting preference model. 
Although being an important insight about the (likely vain) hope of determining a single correct recommendation in many complex decision situations (about which more below), this answer does not address the problem that some principles of decision making may be wrong guides in some situations, possibly including some principles that are commonly applied in the field, hence the need of principled ways of discriminating them.

The prescriptive approach to decision making \citep{bell_decision_1988} has been proposed (in supplement to the normative one and the descriptive one), which insists on incorporating the findings of psychologists about decision maker’s weaknesses of reasoning in the procedures leading to recommendations. I do not discuss this approach specifically here, because its methods can be considered to fall into the two categories I have described here above when it aims at obtaining reflective preference models. When considering normative economics broadly, as I do here, as the methods aiming for valid prescription for an individual or a group, prescriptive approaches are a part of normative economics.

\TODO{Argumentation theory inspired; but they consider arguments are given. Pursues another article of mine.}

\subsection{This approach}
The proposal of this article to capture reflective preferences while avoiding the just described problems is to define normative theories on the basis of a low demanding axiom complemented with principles of reasonings; that allow decision makers to “reconstruct” the reasoning proposed by the principles so as to be able to check that they apply to their situation; therefore permitting to falsify the theories using observable data representing the rejection by the decision makers of these principles. Such theories are therefore both normative and empirical: normative, in the sense of claiming to capture reflective preferences, which provide a better basis for recommendations or policy making than shallow preferences; and empirical, in the sense of the principles of reasoning being exposed to empirical falsification.

@@

One long-term goal of this research program is to give individuals the means to reject theories that they find unconvincing after due consideration of all arguments; and accordingly, give incentives to decision theorists to build defenses of these theories able to convince individuals whose preferences are modeled of the soundness of these theories. This might match \possessivecite{hausman_debate_2010} suggestion that we should use rational persuasion rather than nudge individuals, arguing that it would promote autonomy.

@@

For a clear separation of concern, this article defines first normative theories, then empirical theories for these normative theories. A normative theory makes normative claims only, which have to be accepted or rejected on the basis of the apparent soundness of the axiom on which it rests. An empirical theory for a given normative theory, by contrast, makes empirical claims, hence, can be falsified by data. What I call here an empirical theory, for simplicity, is not however purely empirical, but can also be considered normative, for two reasons. First, its conclusions inevitably rest on the normative validity of the normative theory for which it is built. Second, it claims to capture preferences that provide an appropriate basis for recommendation. Empirical theories of \ac{DJ} could thus be called empirical normative theories of \ac{DJ}.

@@

Thanks to these operators, an empirical theory may adjust its argumentation to each individual. Instead of having to give a single decisive argument in support of its claims, the theory may react to exactly the counter-arguments that a given individual thinks relevant. This task is easier, but still permits to obtain what is desired in this article. First, the claim of the theory, of being able to argue convincingly, can be checked by ensuring that no empirical counter-examples exist, thus, it is a fully empirical claim (in the sense that it requires no axiom). Second, if indeed a theory is convincing in that sense, then it is valid, meaning that the proposition that it claims are \acp{DJ} are indeed \acp{DJ}.

@@

\subsection{Falsifiability}
Given $\args \subseteq \allargs$, $\hist \in \allhist$, and letting $l$ denote the length of $\hist$,
the set $\args \cap \hist(\intvl{1, l})$ denotes the arguments that are both in $\args$ and contained in the sequence $\hist$.
Abusing notation, I write $\args \cap \hist$ to designate $\args \cap \hist(\intvl{1, l})$, therefore using simply $\hist$ in that expression to denote the \emph{range} of $\hist$.

@@

\begin{theorem}
	An empirical theory is convincing iff it has no falsification instances.
\end{theorem}
\begin{proof}
	For all $i \in I, \prop \in \gpropse, \ar \in \gargs$: $\ar$ does not $i$-defend $\prop$ iff $\exists \hist \in \allhist \suchthat \left[\gbeatsinv(\ar) \cap \hist = \emptyset \land \ar \in \histend \land \hist \nileadstost \prop\right]$ iff $\exists \hist \in \allhist$ such that $(i, \prop, \hist, \ar)$ falsifies $\gamma$. Therefore, $\gamma$ is not convincing iff $\exists (i, \prop, \ar) \suchthat \ar$ does not $i$-defend $\prop$ iff $\exists \hist \suchthat (i, \prop, \hist, \ar)$ falsifies $\gamma$ iff $\gamma$ has a falsification instance.
\end{proof}

@@

\begin{theorem}
	A normatively adequate and convincing empirical theory is valid.
\end{theorem}
\begin{proof}
	Given any $i \in I$ and $\prop \in \gpropse$, suffices to show that for some $\ar \in \allargs$, $\ar \ileadstost \prop$; the conclusion then follows thanks to normative adequacy.

@@

%Given a normative theory and two empirical theories for that normative theory and $i \in I$, 
%a \emph{test procedure with $i$} is a finite sequence 
\commentOC{Need the test procedure to be a tree. Atom: $(k, \prop)$ meaning whether ${\hist}^k \ileadstost \prop$. Branch label: atoms linked using or and and operators. From a node $n = (\hist, \ar, \prop)$, let branches depart, such that their labels do not overlap. From a node, if no branch label is true, the node must falsify a theory. In the case of the procedure below, it is not a list because we may query several times the same theory or switch to the other one, depending on the last answer. But if simplifying the theory (single-try), we may then simplify the definition of a test procedure.}

@@

\section{Conclusion}
Back To Bentham? Explorations Of Experienced Utility, Kahneman Wakker, Sarin, propose to measure hedonimeter total for normative.

\citet[p.\ 100]{raiffa_back_1985} (about SEU): “Give therapy to deviators”.
\citet[p.\ 108]{raiffa_back_1985}: “In experiments with my students, my ingenious protestations did not always prove to be ingenious enough.
Yes, they were dazzled (and perhaps confused) but many remained unconvinced; if the chips were down, they would behave just as before.”

@@

\hbadness 10000
\bibliography{simple}
\end{document}

>>>>

\section{Motivation} 
It has escaped nobody’s attention that 
studying reflective preferences, in supplement to shallow preferences, is an important endeavor for economics: we want to recommend to individual decision makers; we want to adopt optimization criteria for society, and this ought not to merely faithfully reproduce unreflectful behavior. This is one important difference between the normative and the (inappropriately called, as will be seen) “descriptive” approach in economics. As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow preferences are not sufficient to inform normative economics.
Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures.

@@

\subsection{Existing approaches for capturing reflective preferences}
\subsubsection{Laundering preferences}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information.

This creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known. As a result, a theory postulating constraints on individual’s behavior on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. 
That this is an important methodological problem is sometimes recognized by authors adopting this approach. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making.

@@

As a consequence of these techniques being not completely formalised and reproducible, each application requires manual intervention of (very skilled) experts. Eliciting reflective preferences therefore requires enormous resources. \Citet[p.\ 572]{pinto_normative_2012} observe that “in many of those surveys that governments use in order to inform social policy there are not enough resources (time and money) in order to reach the stage where preferences are not labile”. There is, to reduce this cost, a need for theories that can be iteratively improved upon and yield reusable knowledge about reflective preferences.

Different analysts, using different methods and different axioms as their favorite conception of rational decision-making, will obtain possibly contradictory conclusions, “elicited” from the decision-makers who hire them. Some analysts admit this possibility and do not consider it to be a problem.
For example, \citet[p.\ 214]{roy_comparison_1995} write: “In our view, these inevitable disagreements do not imply that decision-aid is useless but simply that a single problem may have several valid responses. Given that two different decision-aid models cannot be implemented in the same decision process, the decision-maker must be conscious of the qualitative choices implied by the different models – often conveying the analysts’ own ethical choices – before coming to personal conclusions on the choice to be made. In this domain, the many different approaches reflect in our view the complexity of the researcher's task much more than a scientific weakness.”
Although being an important insight about the (likely vain) hope of determining a single correct recommendation in many complex decision situations (about which more below), this answer does not address the problem that some principles of decision making may be wrong guides in some situations, possibly including some principles that are commonly applied in the field, hence the need of principled ways of discriminating them.

\subsection{Some other related approaches}
The prescriptive approach to decision making \citep{bell_decision_1988} has been proposed (in supplement to the normative one and the descriptive one), which insists on incorporating the findings of psychologists about decision maker’s weaknesses of reasoning in the procedures leading to recommendations. I do not discuss this approach specifically here, because its methods can be considered to fall into the two categories I have described here above when it aims at obtaining reflective preference models. When considering normative economics broadly, as I do here, as the methods aiming for valid prescription for an individual or a group, prescriptive approaches are a part of normative economics.

Some authors have proposed to ground normative theories on observable aspects of nature and individuals, to solve some of the problems mentioned above; for example, \citet{kahneman_back_1997} propose to measure “experienced utility”, considered as a hedonic quality. To the best of my knowledge, these proposals never aim at observing reflective preferences as considered here and thus fall outside the scope of this article.

The field of formal argumentation theory has developed technical approaches to deal with arguments. Such endeavor starts with an “attack” relation, which indicates which arguments argue against which. Various criteria can then be mathematically defined, in the form of a function selecting subsets of arguments given an attack relation, typically representing sets of arguments that “best resist” their counter-arguments. This field has greatly inspired this proposal, in particular, the notion of attack and of reinstatement. This field however does not view the attack relation as observable, but considers it as given, and its methodology is therefore not empirical in the sense seeked for here. (\Citet{cailloux_formal_2019} elaborates on this difference of objective.)

An important literature exists about “debiasing techniques”: techniques that aim at encouraging individuals to think more about their decision problem in order to reduce occurrence of behavioral “mistakes”. Examples include suggesting the individuals to consider alternative points of view, or asking them to give reasons for their choices (an insightful article by \citet{mitchell_libertarian_2004} cite many such studies). This proposal is situated at another level, which could be considered as a way of justifying and comparing such debiasing techniques with no need of a prior agreement on what constitutes a behavioral mistake.

The current proposal pursues a path started with \citet{cailloux_formal_2019}, an article that defines the notion of \ac{DJ} and models that enable interactions with individuals whose \acp{DJ} are to be captured. That article however requires as fundamental observable a “trump” relation, indicating whether an individual thinks that an argument is against another argument; and a “support” relation, indicating whether an individual thinks that an argument supports a proposition. The current proposal defines every required concepts, such as the notion of arguments attacking other arguments, on the basis of a single observable that is compatible with the revealed preference approach priviledged in economic theory, in the sense that it can be defined by reference to what individuals effectively choose under various circumstances. The current proposal also defines explicitly (in the form of an axiom) sufficient requirements to grant normative power to observations, an aspect which was previously left unclear. Finally, the current proposal links \ac{DJ} and deliberated preferences.

\Citet{meinard_justification_2020} discuss some requirements on justification of recommendations, which gives further arguments in favor of studying \acp{DJ}.

\subsection{This approach}
The proposal of this article to capture reflective preferences while avoiding the just described problems is to define normative theories on the basis of a low demanding axiom complemented with principles of reasonings; that allow decision makers to “reconstruct” the reasoning proposed by the principles so as to be able to check that they apply to their situation; therefore permitting to falsify the theories using observable data representing the rejection by the decision makers of these principles. Such theories are therefore both normative and empirical: normative, in the sense of claiming to capture reflective preferences, which provide a better basis for recommendations or policy making than shallow preferences; and empirical, in the sense of the principles of reasoning being exposed to empirical tests.

@@

One long-term goal of this research program is to give individuals the means to reject theories that they find unconvincing after due consideration of all arguments; and accordingly, give incentives to decision theorists to build defenses of these theories able to convince individuals whose preferences are modeled of the soundness of these theories. This might match \possessivecite{hausman_debate_2010} suggestion that we should use rational persuasion rather than nudge individuals, arguing that it would promote autonomy.
It also follows \possessivecite[p.\ 72]{guala_logic_2000} suggestion (discussing the Allais paradox) about obtaining relevant data to compare normative behavioral theories: “Before asking whether agents want to revise their choices or not, \emph{both} rival theories should be proposed” (emphasis in the original text).

@@

For a clear separation of concern, this article defines first normative theories, then empirical theories for these normative theories. A normative theory makes normative claims only, which have to be accepted or rejected on the basis of the apparent soundness of the axiom on which it rests. An empirical theory for a given normative theory, by contrast, makes empirical claims, hence, can be confronted to observations. What I call here an empirical theory, for simplicity, is not however purely empirical, but can also be considered normative, for two reasons. First, its conclusions inevitably rest on the normative validity of the normative theory for which it is built. Second, it claims to capture preferences that provide an appropriate basis for recommendation. Empirical theories of \ac{DJ} could thus be called empirical normative theories of \ac{DJ}.

@@

Thanks to these operators, an empirical theory may adjust its argumentation to each individual. Instead of having to give a single decisive argument in support of its claims, the theory may react to exactly the counter-arguments that a given individual thinks relevant. This task is easier, but still permits to obtain what is desired in this article: there exists a purely empirical property of a theory, called convincingness, that, together with an axiom about the normative adequacy of the observation protocol, implies the correctness of the claims of the theory. In more details, first, the claim of the theory, of being able to argue convincingly, can be checked by ensuring that no falsification instances exist, thus, it is a fully empirical claim (in the sense that it requires no axiom). Second, granted normative adequacy, if indeed a theory is convincing, then it is valid, meaning that the proposition that it claims are \acp{DJ} are indeed \acp{DJ}. These claims are proven in \cref{th:convnofals,th:valid}.

@@

\subsection{Empirical tests}
Given $\args \subseteq \allargs$, $\hist \in \allhist$, and letting $l$ denote the length of $\hist$,
the set $\args \cap \hist(\intvl{1, l})$ denotes the arguments that are both in $\args$ and contained in the sequence $\hist$.
Abusing notation, I write $\args \cap \hist$ to designate $\args \cap \hist(\intvl{1, l})$, therefore using simply $\hist$ in that expression to denote the \emph{range} of $\hist$.

@@

\begin{theorem}
	\label{th:convnofals}
	An empirical theory is convincing iff it has no falsification instances.
\end{theorem}
\begin{proof}
	For all $i \in I, \prop \in \gpropse, \ar \in \gargs$: $\ar$ does not $i$-defend $\prop$ iff $\exists \hist \in \allhist \suchthat \left[\gbeatsinv(\ar) \cap \hist = \emptyset \land \ar \in \histend \land \hist \nileadstost \prop\right]$ iff $\exists \hist \in \allhist$ such that $(i, \prop, \hist, \ar)$ falsifies $\gamma$. Therefore, $\gamma$ is not convincing iff $\exists (i, \prop, \ar) \suchthat \ar$ does not $i$-defend $\prop$ iff $\exists \hist \suchthat (i, \prop, \hist, \ar)$ falsifies $\gamma$ iff $\gamma$ has a falsification instance.
\end{proof}

@@

\begin{theorem}
	\label{th:valid}
	A normatively adequate and convincing empirical theory is valid.
\end{theorem}
\begin{proof}
	Given any $i \in I$ and $\prop \in \gpropse$, suffices to show that for some $\ar \in \allargs$, $\ar \ileadstost \prop$; the conclusion then follows thanks to normative adequacy.

@@

%Given a normative theory and two empirical theories for that normative theory and $i \in I$, 
%a \emph{test procedure with $i$} is a finite sequence 
%\commentOC{Need the test procedure to be a tree. Atom: $(k, \prop)$ meaning whether ${\hist}^k \ileadstost \prop$. Branch label: atoms linked using or and and operators. From a node $n = (\hist, \ar, \prop)$, let branches depart, such that their labels do not overlap. From a node, if no branch label is true, the node must falsify a theory. In the case of the procedure below, it is not a list because we may query several times the same theory or switch to the other one, depending on the last answer. But if simplifying the theory (single-try), we may then simplify the definition of a test procedure.}

@@

\section{Conclusion}
(To be written.)

- A theory will never be absolutely convincing in the sense defined here. \citet[p.\ 100]{raiffa_back_1985} (about SEU): “Give therapy to deviators”. \citet[p.\ 108]{raiffa_back_1985}: “In experiments with my students, my ingenious protestations did not always prove to be ingenious enough.
Yes, they were dazzled (and perhaps confused) but many remained unconvinced; if the chips were down, they would behave just as before.”
Existence of a falsification instance does not mean that a theory should be rejected. Instead, comparing theories in the suggested way permits to gather information about their relative convincingness power, which in turn will inform on the appropriateness of using them normatively. Supplementary criteria might be required to decide on normative power. For example, policies that would contradict “very strongly” shallow preferences could be considered as excluded, for practical or for principled reasons, even possibly in situations where it could be shown convincingly that such policies would better satisfy reflective preferences.
Relatedly, it is not claimed here that what society ought to do is exactly what \acp{DJ} indicate, only that reflective preferences provide an appropriate basis for such normative endeavors, in supplement to other criteria. Further discussions are required, which hopefully the formal definitions here can help clarify.

- Theories must be able to adapt their claims to individuals. The mechanism described here that adapts arguments to individuals can be adapted to such goal. This is left for future work.

- Back to the claim that a decision problem may admit several valid responses. The current approach partly permits to develop theories that agree with this claim, because these theories may be incomplete (in the sense of defining only a partial deliberated preference). Taking this view seriously, however, requires to account for meta-preferences (cf. Sen): a (possibly partial) ordering on the possible definitions of one’s preference. As an example, a theory should be able to express that $i$ will adopt either the view that $a > b > c$, or the view that $c > b > a$, but not $a > c > b$.
%Relatedly, akrasia might be considered as well. An individual may be able to declare that she would prefer having a preference rather than another one; or be put in a situation that helps her follow a preference rather than another one. As an example, an individual might be willing to declare

\bibliography{simple}
\end{document}

@@
